Replicator-mutator and noise-perturbed replicator dynamics

### 2014-09-16 paper/02_noisyRMD.tex
improved structure Sec 02; data for convex vs. non-convex strats for small state spaces

\section{Noise-perturbed replicator dynamic}

\subsection{Preliminaries}

\paragraph{Notation.} Fix a signaling game with states $\States$,
messages $\Messgs$ and acts $\Acts$. % Let $\Util_{\sen,\rec} \mycolon
% \States \times \Messgs \times \Acts \rightarrow \mathds{R}$ be the
% player's utility function.

Pure sender (receiver) strategies are functions $\Spure \in
\Messgs^\States$ ($\Rpure \in \Acts^\Messgs$). Mixed sender (receiver)
strategies are functions $\Smixed \in \Delta(\Messgs^\States)$
($\Rmixed \in (\Acts^\Messgs)$). The latter give the relative
population frequencies of the former. We write $\Smixed_i$ for the
frequency $\Smixed(\Spure_i)$ of pure strategy $\Spure_i$ (also for
the receiver).

Let the sender's (receiver's) behavioral strategies be functions
$\Sstrat \in \Delta(\Messgs)^\States$ ($\Rstrat \in
\Delta(\Acts)^\Messgs$). Every mixed strategy $\Smixed$ converts to a
unique behavioral strategy defined by:
\begin{align*}
  \Sstrat(\messg \probbar \state) = \sum_{\Spure(\state) = \messg} \Smixed(\Spure)\,.
\end{align*} 
Let $G$ be this mapping from mixed to behavioral strategies. Notice
that $G$ is \emph{not} an injection, as many mixed strategies map onto
the same behavioral strategy.

For behavioral strategies, let $\EU(\messg, \state, \Rstrat)$ and
$\EU(\act, \messg, \Sstrat)$ be the players' expected utilities of
each action in each choice point as usual. For mixed strategies we
define the players' fitness in the usual manner. Let $F_i^{\Rmixed}$
be $\Spure_i$'s fitness given $\Rmixed$ and $F_i^{\Smixed}$ be
$\Rpure_i$'s fitness given $\Smixed$. Then $\Phi(\Smixed,\Rmixed) =
\sum_{k} \Smixed_k \cdot F_k^{\Rmixed}$ is the average fitness in the
sender population and $\Phi(\Rmixed,\Smixed) = \sum_{k} \Rmixed_k
\cdot F_k^{\Smixed}$ the average fitness in the receiver population.

>>>>

\section{Replicator diffusion dynamic}

The replicator diffusion dynamic (\rdd) was first introduced by
\citet{Correia2013:The-Bivalent-Tr} as a noise-perturbed variant of
the replicator dynamic (\rd) in behavioral strategies. We begin by
recapitulating the formulation of the \rd in behavioral
strategies. Then we define the replicator diffusion dynamics in
behavioral strategies, following \citet{Correia2013:The-Bivalent-Tr}.
Finally we show that the \rdd is a special case of the \rmd.

\subsection{Replicator dynamic in behavioral strategies}

Fix a signaling game with states $\States$, messages $\Messgs$ and
acts $\Acts$. Let $\Pr(\cdot) \in \Delta{\States}$ be the prior
distribution over states and $\Util_{\sen,\rec} \mycolon \States
\times \Messgs \times \Acts \rightarrow \mathds{R}$ the senders's and
receiver's utility functions. The sender's behavioral strategies are
functions $\Sstrat \in \Delta(\Messgs)^\States$; the receiver's are
functions $\Rstrat \in \Delta(\Acts)^\Messgs$. Define the expected
utility of actions in each choice point are defined as usual:
\begin{align*}
  \EU(\messg \myts ; \myts \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act \myts ; \myts \messg, \Sstrat) & = \sum_{\state \in
    \States} \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg \myts ; \myts\state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act \myts ; \myts \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$ in
such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:\todo{could be shortened by using
  $\propto$ notation}
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state_i) & = \frac{\Sstrat(\messg \probbar \state_j) \cdot
    \EU(\messg,\state_j,\Rstrat)} {\Phi(\state_j,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\state_i \probbar \messg) & = \frac{\Rstrat(\state_j \probbar \messg) \cdot
    \EU(\state_j,\messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}

\todo[inline]{example? sim-max game 10 states, after so and so many rounds?}

\subsection{Replicator diffusion dynamic in behavioral strategies}

The replicator diffusion dynamic adds probabilistic confusion of
similar states to the replicator dynamics. Fix a sim-max game with
$\States = \Acts$ and a confusion matrix $C$. $C$ is a row-stochastic
matrix whose elements $C_{ij}$ give the probability that $\state_i$ is
realized as $\state_j$. The confusability of states affects senders
and receivers alike, but in slightly different ways (see
Figure~\ref{fig:noise-perturbation-of-strategies}). For the sender,
$C_{ij}$ is the probability that the actual states \mystate{i} is
perceived as \mystate{j}. For the receiver, $C_{ij}$ is the
probability that \mystate{j} is the interpretation that is actually
formed when \mystate{i} is the output of the receiver's strategy.

\begin{figure}
  \centering

    \begin{tikzpicture}[node distance = 2cm, thick]

      \begin{scope}

      \node[rectangle, draw=mycol, fill=mycol!25, thick] (actual)
      {actual state $\mystate{i}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      actual] (perceived) {perceived state $\mystate{j}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      perceived] (output) {chosen message $\messg$};

      \node[rectangle, thick, below of = output, node distance=1cm]
      (sender) {sender};

      \draw[->] (actual) -> (perceived) node[midway,left] {noise
        $C_{ij}$};

      \draw[->] (perceived) -> (output) node[midway,left] (label)
      {strategy $\Sstrat(\messg \probbar \mystate{j})$};

      \begin{pgfonlayer}{background}
        \node [draw=black!50, fill=black!20,fit=(actual) (label)
        (output)] {};
      \end{pgfonlayer}
    \end{scope}

      \begin{scope}[xshift=5cm]

      \node[rectangle, draw=mycol, fill=mycol!25, thick] (message)
      {observed message $\messg$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      message] (target) {target interpretation $\mystate{i}$};

      \node[rectangle, draw=mycol, fill=mycol!25, thick, below of =
      target] (realized) {realized interpretation $\mystate{j}$};

      \node[rectangle, thick, below of = realized, node distance=1cm]
      (sender) {receiver};

      \draw[->] (message) -> (target) node[midway,right] (bla) {strategy
        $\Rstrat(\mystate{i} \probbar \messg)$};

      \draw[->] (target) -> (realized) node[midway,right]  {noise $C_{ij}$};

      \begin{pgfonlayer}{background}
        \node [draw=black!50, fill=black!20,fit = (message) 
        (realized) (bla)] {};
      \end{pgfonlayer}
    \end{scope}

  \end{tikzpicture}

  \caption{Effect of confusion of states on sender and receiver
    choices.}
  \label{fig:noise-perturbation-of-strategies}
\end{figure}

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ on their noise-perturbed realizations
$C(\Sstrat)$ and $C(\Rstrat)$. The idea is that $\Sstrat$ and
$\Rstrat$ are what, on average, the idealized, confusion-free behavior
would be, while $C(\Sstrat)$ and $C(\Rstrat)$ are behavioral
strategies that describe the agents' actual noise-perturbed
probabilistic behavior. If we conceive of behavioral strategies
$\Sstrat$ and $\Rstrat$ as row-stochastic matrices, the effect of
state confusability is easily captured by matrix multiplication as:
\begin{align*}
  C(\Sstrat) & = C \Sstrat &    C(\Rstrat) & = \Rstrat C\,.
\end{align*}

The discrete-time replicator diffusion dynamic takes the replicator
dynamic as basic, but factors in the confusion of states at each
update step in a sequential update:
\begin{align*}
  \RDD(\Sstrat) & = C(\RD(\Sstrat)) &   \RDD(\Rstrat) & = C(\RD(\Rstrat)) 
\end{align*}
This is equivalent to the following definition, given by
\citet{Correia2013:The-Bivalent-Tr}:
\begin{align*}
  \RDD(\Sstrat)(\messg \probbar \state_i) & = \sum_{j} C_{ij} \cdot
  \frac{\Sstrat(\messg \probbar \state_j) \cdot
    \EU(\messg,\state_j,\Rstrat)} {\Phi(\state_j,\Rstrat)} \\
    \RDD(\Rstrat)(\state_i \probbar \messg) & = \sum_{j} C_{ij} \cdot
  \frac{\Rstrat(\state_j \probbar \messg) \cdot
    \EU(\state_j,\messg,\Sstrat)} {\Phi(\messg,\Sstrat)}  \,.
\end{align*}

The idea behind the sequential definition of the \rdd is that, at each
time step, strategies are gradually optimized along the current
fitness landscape, as described by the \rd, but the realization of
optimized strategies is bound to be noisy, due to confusion of
states. It is obvious that other (discrete-time) evolutionary dynamics
can be subjected to state-confusability in an analogous fashion
\citep{Correia2013:The-Bivalent-Tr}. In the case of the \rd, however,
perturbation by state-confusion has a prominent close relative in the
replicator mutator dynamic.

\subsection{Relation with the replicator mutator dynamic}

The replicator mutator dynamic (\rmd) has been proposed by Martin
Nowak first in the context of signaling game models for the evolution
of grammar
\citep[e.g.][]{KomarovaNiyogi2001:The-Evolutionar,NowakKomarova2001:Evolution-of-Un,Nowak2006:Evolutionary-Dy}. The
relation of the \rmd to other prominent evolutionary dynamics is well
understood \citep{PageNowak2002:Unifying-Evolut}. The \rmd has seen
further fruitful applications in the context of signaling games
\citep{HutteggerSkyrms2010:Evolutionary-Dy}. It is therefore desirable
to relate the \rdd that we propose here to the \rmd.

A direct comparison between \rdd and the \rmd is not possible, because
the usual formulation of the latter is in its continuous-time form,
and based on mixed strategies, not behavioral strategies. We therefore
give a discrete-time formulation of the \rmd in behavioral strategies,
which is independently useful. The relation between \rdd and \rmd will
then be plain to see.

\subsubsection{Replicator mutator dynamic}

Pure sender (receiver) strategies are functions $\Spure \in
\Messgs^\States$ ($\Rpure \in \Acts^\Messgs$). Mixed sender (receiver)
strategies are functions $\Smixed \in \Delta(\Messgs^\States)$
($\Rmixed \in (\Acts^\Messgs)$). The latter give the relative
population frequencies of the former. We write $\Smixed_i$ for the
frequency $\Smixed(\Spure_i)$ of pure strategy $\Spure_i$ (also for
the receiver). For mixed strategies, we define the players' fitness in
the usual manner. Let $F_i^{\Rmixed}$ be $\Spure_i$'s fitness given
$\Rmixed$ and $F_i^{\Smixed}$ be $\Rpure_i$'s fitness given
$\Smixed$. Then $\Phi(\Smixed,\Rmixed) = \sum_{k} \Smixed_k \cdot
F_k^{\Rmixed}$ is the average fitness in the sender population and
$\Phi(\Rmixed,\Smixed) = \sum_{k} \Rmixed_k \cdot F_k^{\Smixed}$ the
average fitness in the receiver population.

Every mixed strategy $\Smixed$ converts to a unique behavioral
strategy defined by:
\begin{align*}
  \Sstrat(\messg \probbar \state) = \sum_{\Spure(\state) = \messg} \Smixed(\Spure)\,.
\end{align*} 
Let $G$ be this mapping from mixed to behavioral strategies. Notice
that $G$ is \emph{not} an injection, as many mixed strategies map onto
the same behavioral strategy. \todo{this should not be here}

### 2014-09-16 snippets/data_exploration/data_exploration_tmp.Rmd
improved structure Sec 02; data for convex vs. non-convex strats for small state spaces

load and pre-process the data

``` {r warnings=FALSE, message=FALSE}

@@

data = read.csv('~/Desktop/data/svn/vagueness-games/results/20140915-190401.csv', 
                colClasses=c('numeric','factor','numeric','numeric','numeric',
                             'factor','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','numeric','numeric',
                             'numeric','character','character'))

ds = data[,c(1,4,7,8,11:16)] # only relevant data
dm = melt(ds,id.vars = c("Number.of.states", "Impairment")) # melted data

@@

pairs(~ Number.of.states + Impairment + Speaker.entropy + Hearer.entropy +
        Speaker.Voronoiness + Hearer.Voronoiness +
        Speaker.informativity + Hearer.informativity + Expected.utility + Iterations, data=ds)
```

@@

pd <- position_dodge(.05)
sp = ggplot(dsub, aes(x=Impairment, y=value, colour=factor(Number.of.states))) + 
     geom_point(position = pd) +
     geom_errorbar(aes(ymin=value-ci, ymax=value+ci), width=0, position=pd) +
     geom_line(position=pd)

>>>>

``` {r warnings=FALSE, message=FALSE, echo=FALSE}

@@

data = read.csv('~/Desktop/data/svn/vagueness-games/results/20140916-133324.csv', 
                colClasses=c('numeric','factor','numeric','numeric','numeric',
                             'factor','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','numeric','numeric',
                             'numeric','factor','factor','character','character'))

ds = data[,c(1,4,7,8,9,10,11:16)] # only relevant data
dm = melt(ds,id.vars = c("Number.of.states", "Impairment")) # melted data

@@

pairs(~ Number.of.states + Impairment + Speaker.entropy + Hearer.entropy + Speaker.convexity +
        Speaker.Voronoiness + Hearer.Voronoiness +
        Speaker.informativity + Hearer.informativity + Expected.utility + Iterations, data=ds)
```

@@

pd <- position_dodge(.01)
sp = ggplot(dsub, aes(x=Impairment, y=value, colour=factor(Number.of.states))) + 
     geom_point(position = pd) +
     geom_errorbar(aes(ymin=value-ci, ymax=value+ci), width=0, position=pd) +
     geom_line(position=pd)
